---
title: "Does Allen's rule still apply to the world's Birds?"
author: "Karim Bolis u6671566"
date: "14/10/23"
output:  
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
---

# Word/figure count

Words: 1989
Figures: 7

# Location on GitHub

https://github.com/karim-bolis/DS4B-final-project

# Data Description

Avonet is an open-source database of morphological, geographical and ecological data for all bird species. It was compiled by Catherine Sheard and Joseph Tobias to serve as a tool for testing out theories in evolutionary biology, ecology and ecosystem science for the birds. (Tobias et al. 2021)

# Questions/Aims

## Introduction

As the climate gets hotter, ornithologists have found shifts in bird sizes, geographic ranges and phenology. (Tian & Benton 2020, Baldwin et al. 2023) This has sparked an interest in predicting future shifts using biogeographical rules which summarise large global trends. (Tian & Benton 2020) Allen's rule, one of the oldest ones, states that in hotter climates, endotherm appendages tend to become larger, as the increase in surface-area-to-volume ratio results in more heat exchange with the environment. (Allen 1877) Allen's rule has been studied extensively, often in conjunction with Bergmann's rule, which links body size with temperature. (Xu 2023, Symonds & Tattersall 2010,  Frohlich et al. 2023, Baldwin et al. 2023) Yet, published literature has only looked into bird beak dimensions and tarsus (foot) length as appendage size representatives, often neglecting wings and tails. The presence of feathering on these appendages may complicate results. During flight, birds lose 99% of the heat generated, mostly through their wings, yet heat conservation is also achieved by holding the wings close to one's torso (Weeks et al. 2023). Similarly, beaks are simultaneously involved in both heat dissipation and cold tolerance (Weeks et al. 2023) This study aims to understand how Wing and Tail Length factor into Allen's rule across the world's species.

# Raw data

Download the file named "AVONET from Supplementary dataset 1.xlsx" from https://figshare.com/articles/dataset/AVONET_morphological_ecological_and_geographical_data_for_all_birds_Tobias_et_al_2021_Ecology_Letters_/16586228.

WorldClim data is downloaded below.

```{r}
library(raster)

worldclim = getData("worldclim",var="bio",res=2.5, path="raw_data")

```

# Data wrangling

```{r}
library(tidyverse)
library(readxl)

raw_avonet = read_excel("raw_data/AVONET Supplementary dataset 1.xlsx", sheet=4)
avonet = as_tibble(raw_avonet)
avonet = avonet %>%
  mutate(row_id = row_number()) %>%
  select(row_id, everything()) %>%
  mutate(Centroid.Latitude=as.double(Centroid.Latitude)) %>%
  mutate(Centroid.Longitude=as.double(Centroid.Longitude)) %>%
  mutate(Hand.Wing.Index=`Hand-Wing.Index`)

write_csv(avonet, file = "processed_data/tidied_data.csv")

excluded_data <- read_csv("processed_data/excluded_data.csv")
tidied_data <- read_csv("processed_data/tidied_data.csv")
cleaned_data <- tidied_data %>%
  anti_join(excluded_data, by = "row_id")

write_csv(cleaned_data, file = "processed_data/cleaned_data.csv")

avonet <- read_csv("processed_data/cleaned_data.csv")

```

# Sanity checks

Most of the sanity checking was done in the data cleaning document. These checks ensure that data types are consistent with how they appear, that all species have location information to find their living range mean annual temperature and that none of the traits studied have been inferred from close relatives.

```{r}

summary(avonet)

any(is.na(avonet$Centroid.Latitude))
any(is.na(avonet$Centroid.Longitude))

unique(avonet$Traits.inferred)

```

# Report

## Obtaining annual temperature

```{r}
library(raster)

source("extra_functions.R")

worldclim <- worldclim[[1]]
plot(worldclim)

avonet = add_mean_annual_temp(avonet)

head(avonet)

```

## PCA
The violin plot shows that most values in the distributions of wing and tail length sit within a small range, with lots of outliers. They have a clear linear relationship with narrow confidence intervals. To quantify the effect size, I found their pearson and spearman correlation coefficients. While both coefficients assume a linear relationship, pearson is heavily affected by outliers, so spearman performs better.

This strong relationship indicates that a Principal component analysis, which reduces data dimensionality, could provide a way to capture both sets of data and provide a stronger correlate with mean annual temperature. This technique was previously used to capture bill size from its 4 dimensional measurments. (Baldwin et al. 2023) The first principal component, capturing 86% of both variables, was added to avonet.

```{r}
library(psych)

summary(avonet$Wing.Length)
summary(avonet$Tail.Length)

ggplot(avonet, aes(x = 1, y = Wing.Length)) +
  geom_violin(fill = "blue") +
  labs(y = "Trait Value") +
  labs(x = "Distributions") +
  geom_violin(data = avonet, aes(x = 2, y = Tail.Length), fill = "red") +
  scale_x_continuous(breaks = c(1, 2), labels = c("Wing Length", "Tail Length"))

ggplot(avonet, aes(x = Wing.Length, y = Tail.Length)) +
  geom_point(size = 0.01) +
  scale_y_log10() +
  geom_smooth(method = 'lm', se=TRUE, level=0.99999999)

correlation_value_pearson = cor(avonet$Wing.Length, avonet$Tail.Length, method = "pearson")
correlation_value_spearman = cor(avonet$Wing.Length, avonet$Tail.Length, method = "spearman")
print(correlation_value_pearson)
print(correlation_value_spearman)

pca_model = principal(select(avonet,Wing.Length,Tail.Length), nfactors = 2, rotate='none')
print(pca_model)

pc1 = pca_model$scores[,1]

avonet$wing_tail_pc1 = pc1

```

## Correlations

I chose to study family-level correlations, as families usually contain a sufficiently large temperature range and number of species, keeping potentially confounding effects of morphology, behavior, and ecology variation low. Additionally, families have vastly different trait value distributions, as seen in the histograms. The families studied were filtered to favor a temperature range and no. of species > 10.

```{r}
Anatidae_Tyrannidae <- avonet %>%
  filter(Family3 %in% c("Anatidae", "Tyrannidae"))

ggplot(Anatidae_Tyrannidae, aes(x = Wing.Length)) +
  geom_histogram(fill = "blue", color = "black") +
  facet_wrap(~Family3, ncol = 2) +
  labs(x = "Wing Length", y = "Frequency") +
  ggtitle("Histograms of PC1 Values for Anatidae and Tyrannidae")

avonet = avonet %>%
  group_by(Family3) %>%
  mutate(temp_range=max(mean_annual_temp)-min(mean_annual_temp))

avonet = filter(avonet,temp_range>10)

avonet = avonet %>%
  group_by(Family3) %>%
  mutate(no_species_in_family=n())

avonet = filter(avonet,no_species_in_family>10)
```

I first found the distribution of p values across families for each trait. I then transformed the results to produce p-value distribution boxplots, which showed that most relationships are not significant for the traits, with only the Hand-Wing Index median falling at or below the 0.05 line, and most traits containing relatively few significant families out of 96 families total.

While the significant relationship counts featured some for all traits, Wing Length, Tail Length, Hand-Wing Index (HWI) and Kipps' Distance contained the most. The PC1 from the Wing/Tail Length PCA seems to show a correlation between the two traits, rather than improving on them.
The HWI is a metric of flight efficiency and ability to disperse in birds (Tobias et al. 2021, Claramunt 2021). It is equivalent to Kipps' length, explaining their similar numbers, and is not a measure of appendage length, more of relative ratios. (Claramunt 2021) Research has suggested that tropical species in hotter climates have less dispersal ability, increasing their sensitivity to habitat disturbance. (Sheard et al. 2020) This may explain why it is the most negatively correlated trait.


```{r fig.width=10}
correlation_values <- avonet %>%
  group_by(Family3) %>%
  summarize(
    p_value_Wing_Length = cor.test(mean_annual_temp, Wing.Length)$p.value,
    r_value_Wing_Length = cor(mean_annual_temp, Wing.Length),
    p_value_Tail_Length = cor.test(mean_annual_temp, Tail.Length)$p.value,
    r_value_Tail_Length = cor(mean_annual_temp, Tail.Length),
    p_value_wing_tail_pc1 = cor.test(mean_annual_temp, wing_tail_pc1)$p.value,
    r_value_wing_tail_pc1 = cor(mean_annual_temp, wing_tail_pc1),
    p_value_Secondary1 = cor.test(mean_annual_temp, Secondary1)$p.value,
    r_value_Secondary1 = cor(mean_annual_temp, Secondary1),
    p_value_Hand_Wing_Index = cor.test(mean_annual_temp, Hand.Wing.Index)$p.value,
    r_value_Hand_Wing_Index = cor(mean_annual_temp, Hand.Wing.Index),
    p_value_Kipps_dist = cor.test(mean_annual_temp, Kipps.Distance)$p.value,
    r_value_Kipps_dist = cor(mean_annual_temp, Kipps.Distance)
  ) %>%
  ungroup()

correlations_p_values = select(correlation_values, Family3, p_value_Wing_Length, p_value_Tail_Length, p_value_wing_tail_pc1, p_value_Secondary1, p_value_Hand_Wing_Index, p_value_Kipps_dist)

correlations_p_values = correlations_p_values %>% pivot_longer(cols=c('p_value_Wing_Length', 'p_value_Tail_Length', 'p_value_wing_tail_pc1', 'p_value_Secondary1', 'p_value_Hand_Wing_Index', 'p_value_Kipps_dist'),
                    names_to='Trait',
                    values_to='p_value')

correlations_p_values <- correlations_p_values %>%
  mutate(Trait = substr(Trait, 9, nchar(Trait)))

correlations_r_values = select(correlation_values, Family3, r_value_Wing_Length, r_value_Tail_Length, r_value_wing_tail_pc1, r_value_Secondary1, r_value_Hand_Wing_Index, r_value_Kipps_dist)

correlations_r_values = correlations_r_values %>% pivot_longer(cols=c('r_value_Wing_Length', 'r_value_Tail_Length', 'r_value_wing_tail_pc1', 'r_value_Secondary1', 'r_value_Hand_Wing_Index', 'r_value_Kipps_dist'),
                    names_to='Trait',
                    values_to='r_value')

correlations_r_values <- correlations_r_values %>%
  mutate(Trait = substr(Trait, 9, nchar(Trait)))

correlations_p_r_values <- inner_join(correlations_p_values, correlations_r_values, by = c("Family3", "Trait"))

ggplot(correlations_p_r_values, aes(x=Trait, y=p_value, color=Trait)) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  geom_boxplot() +
  scale_x_discrete(labels = c("Hand-Wing Index","Kipps Distance","Secondary Length","Tail Length","Wing Length","Wing Tail PC1")) +
  theme(axis.text.x = element_text(size = 10)) +
  labs(x = "Trait p-values", y = "Distributions", title = "Distribution of P values of trait/temperature correlations across families")

length(unique(avonet$Family3))
nrow(filter(correlation_values,p_value_Wing_Length<0.05))
nrow(filter(correlation_values,p_value_Tail_Length<0.05))
nrow(filter(correlation_values,p_value_wing_tail_pc1<0.05))
nrow(filter(correlation_values,p_value_Secondary1<0.05))
nrow(filter(correlation_values,p_value_Hand_Wing_Index<0.05))
nrow(filter(correlation_values,p_value_Kipps_dist<0.05))

```

The violin plots show that while most significantly correlated families for a trait show a negative correlation, all traits have at least some families where a positive trend is seen. Interestingly, correlation coefficients become scarce around 0, indicating that heavy selection pressures apply on each trait, regardless of direction.
Assessing the families with the most correlation (by correlation coefficient) shows the  Remizidae family has the largest negative trend for Wing Length and Kipps' Distance, explaining 61-62% of the trait's variation through the mean annual temperature (r=-0.78^2 --> R=0.61). Additionally, it is amongst the largest negatively correlated families for Tail Length. Amongst traits which directly measure appendage size, the Gruidae family shows the strongest positive correlation for Tail Length.

Frohlich et al. found that various body/appendage size configurations allow organisms to adapt to climates. (Frohlich et al. 2023) In their findings, tarsus size actually decreased with temperature in smaller birds, producing the opposite effect to Allen's rule, while larger birds were more likely to obey the rule. (Frohlich et al. 2023) Indeed, Allen himself hypothesised that larger birds would be more likely to follow his rule. (Frohlich et al. 2023) The vastly different masses of Remizidae and Gruidae supports this. Additionally, Xu et al. found that passerines (Remizidae) exhibit a negative trend with annual mean temperature, while non-passerines (Gruidae) exhibit a positive trend for bill and tarsus length. (Xu 2023) It is possible, then, that Wing and Tai length are under similar selection pressures to that producing these trends in tarsus length.

```{r fig.width=10}

correlations_significant = filter(correlations_p_r_values,p_value<0.05)

ggplot(correlations_significant, aes(x=Trait, y=r_value, color=Trait)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_violin() +
  geom_jitter(size=0.5, alpha=0.5, width = 0.1) +
  theme_minimal() +
  scale_x_discrete(labels = c("Hand-Wing Index","Kipps Distance","Secondary Length","Tail Length","Wing Length","Wing Tail PC1")) +
  theme(axis.text.x = element_text(size = 10)) +
  labs(x = "Trait r values", y = "Distributions", title = "Distribution of r values of trait/temperature correlations across families")

extreme_family_finder <- function(correlations,trait) {
  trait_correlations = filter(correlations_significant,Trait==trait)

  most_negative = trait_correlations %>%
    arrange(r_value) %>%
    slice(1)

  most_positive = trait_correlations %>%
    arrange(desc(r_value)) %>%
    slice(1)

  print(most_negative)
  print(most_positive)
}

extreme_family_finder(correlations_significant,"Wing_Length")

extreme_family_finder(correlations_significant,"Kipps_dist")

extreme_family_finder(correlations_significant,"Tail_Length")

extreme_family_finder(correlations_significant,"Hand_Wing_Index")

extreme_family_finder(correlations_significant,"Secondary1")

extreme_family_finder(correlations_significant,"wing_tail_pc1")


print(arrange(filter(correlations_p_r_values,Trait=="Tail_Length"),r_value))

mean(filter(avonet,Family3=="Remizidae")$Mass)
mean(filter(avonet,Family3=="Gruidae")$Mass)

```


This theory was examined in the plot below, where r values for a given trait are plotted against the corresponding mean mass for the same family. While the minimum mass value has a lower r value than the maximum for most traits, a clear positive trend is not seen in any of the traits. This indicates the likely presence of various confounding effects, reflecting the various dimensions and environmental selection pressures on Wing and Tail dimensions.


```{r}

mean_mass_per_family <- avonet %>%
  group_by(Family3) %>%
  summarize(mean_mass = mean(Mass, na.rm = TRUE))

correlations_significant <- correlations_significant %>%
  left_join(mean_mass_per_family, by = "Family3")

ggplot(correlations_significant, aes(x = mean_mass, y = r_value, colour = Trait)) +
  geom_point() +
  geom_line() +
  scale_x_log10() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ Trait) +
  labs(x = "X-axis", y = "Y-axis")
  theme_minimal()
  
```




# References

Allen, JA 1907, ‘The Influence of Physical Conditions in the Genesis of Species’, Scientific American, vol. 63, no. 1636supp, pp. 26217–26219.

Baldwin, JW, Garcia‐Porta, J & Botero, CA 2023, ‘Complementarity in Allen’s and Bergmann’s rules among birds’, Nature Communications, vol. 14, Nature Portfolio, no. 1.

Claramunt, S 2021, ‘Flight efficiency explains differences in natal dispersal distances in birds’, Ecology, vol. 102, no. 9.

Fick, S.E. and R.J. Hijmans, 2017, 'WorldClim 2: new 1km spatial resolution climate surfaces for global land areas', International Journal of Climatology, vol. 37, no. 12, pp. 4302-4315.

Frӧhlich, A, Kotowska, D, Martyka, R & Symonds, MRE 2023, ‘Allometry reveals trade-offs between Bergmann’s and Allen’s rules, and different avian adaptive strategies for thermoregulation’, Nature Communications, vol. 14, no. 1.

Sheard, C, Neate-Clegg, MHC, Alioravainen, N, Jones, SEI, Vincent, C, MacGregor, HEA, Bregman, TP, Claramunt, S & Tobias, JA 2020, ‘Ecological drivers of global gradients in avian dispersal inferred from wing morphology’, Nature Communications, vol. 11, no. 1.

Symonds, Matthew R  E. & Tattersall, Glenn J 2010, ‘Geographical Variation in Bill Size across Bird Species Provides Evidence for Allen’s Rule’, The American Naturalist, vol. 176, no. 2, pp. 188–197.

Tian, L & Benton, MJ 2020, ‘Predicting biotic responses to future climate warming with classic ecogeographic rules’, Current Biology, vol. 30, no. 13, pp. R744–R749.

Tobias, JA, Sheard, C, Pigot, AL, Devenish, AJM, Yang, J, Sayol, F, Neate‐Clegg, MHC, Alioravainen, N, Weeks, TL, Barber, RA, Walkden, PA, ... Claramunt, S & Darski, B 2022, ‘AVONET: morphological, ecological and geographical data for all birds’, in T Coulson (ed.), Ecology Letters, vol. 25, no. 3, pp. 581–597.

Weeks, B., Harvey, C., Tobias, J.A., Sheard, C., Zhou, Z., Fouhey, D.F., 2023, 'Skeletal morphology of bird wings is determined by thermoregulatory demand for heat dissipation in warmer climates', In Manuscript.

Xu, Y, Price, ML, Que, P, Zhang, K, Shang, S, He, X, Wen, Z & Wang, B 2023, ‘Ecological predictors of interspecific variation in bird bill and leg lengths on a global scale’, Proceedings of The Royal Society B: Biological Sciences, vol. 290, Royal Society, no. 2003.
